---
title: "Plotting Overview and Refresher"
format: html
---


Resources:

- [Data Carpentry - Data Analysis and Visualisation for Ecologists](https://datacarpentry.github.io/R-ecology-lesson/visualizing-ggplot.html)
- [R for Data Science - Visualisation](https://r4ds.hadley.nz/data-visualize.html)



## Refresher


A more in depth intro to ggplot2 is covered in [Data Carpentry - Data Analysis and Visualisation for Ecologists](https://datacarpentry.github.io/R-ecology-lesson/visualizing-ggplot.html). This is a brief overview/refresher.


### Setup

```{r}
#| message: false

library(tidyverse)
library(palmerpenguins)

# brings 'penguins' dataset into name space
attach(penguins)
# brings 'penguins_raw' dataset into name space
attach(penguins_raw)
```

### Basics of Plotting with {ggplot2}

The basic premise of ggplot is that there are 3 main components to a plot:

-   the *data*
-   a mapping of the data to *aesthetics* of the plot e.g. x or y axes
    -   this can be done for all layers of a plot at once by including it within the `aes()` within the `ggplot()` function
    -   or to specific layers by using `aes()` within a *geom* function
-   a *geometric* representation of the data e.g points, bars, boxplot

and the plots are built in layers

```{r}
ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g,
                y = flipper_length_mm)
) +
  geom_point()
```

And we can modify the properties of layer to a specific value

```{r}

ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g,
                y = flipper_length_mm)
) +
  geom_point(colour = "blue")
```

Or the property can be mapped back to the data - when this is the case we use `aes()`

```{r}

ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g,
                y = flipper_length_mm)
) +
  geom_point(mapping = aes(colour = island))
```

Plots can also be built incrementally by assignment

```{r}
penguin_plot <- ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g,
                y = flipper_length_mm,
                colour = island)
) 

penguin_plot <- penguin_plot +
  geom_point()

penguin_plot
```