---
title: "Beyond Intro R"
format: html
editor: visual
---


The goal of this workshop is to build upon "introductory content" that is found in other workshops such as [Data Carpentry - Data Analysis and Visualisation for Ecologists](https://datacarpentry.github.io/R-ecology-lesson/).

_The workshop itself is a bit piecemeal and covers a range of topics and is mostly designed as a source of quick reference for the instructor during the session_

The key themes of the workshop are to:

- Cover more about plot customisation using ggplot2
- Look at some of the useful helper functions for data wrangling that exist (mostly in the tidyverse)
- Introduce building and applying custom functions
- And some setup/configuration settings that can be useful


The main reference material is the package documentation itself

## R Syntax Refresher


### Maths

Mathematical operations:

- `+` and `-` for addition and subtraction
- `*`, `/`, and `%%` for multiplication, division, and remainder division
- `^` for exponentiation
- `(` and `)` for bracketing
- Precedence of operators follows BEDMAS
```{r}
1 + 6
5 - 3
2 * 9
8 / 4
9 %% 2
2 ^ 3
```

Assignment

```{r}
a <- 7

a
```

Vectors

- must be all of the same type
    - numeric
    - character
    - boolean
```{r}
b <-c(2, 5, 6)
```

Indexing (starts from 1)
```{r}
# 3rd item
b[3]

# Exclude the 2nd item
b[-2]
```

Subsetting with by comparison

```{r}
# items in b larger than 3
b[ b > 3]

# items in b that are equal to 6
b[ b == 6]
```

