---
title: "R Configuration"
format: html
---

## Configuration

As part setting R up to work best for us, we're also wanting to make things easier on ourselves to that we can be more productive during the "do stuff" stage.

There is a hidden file called .Rprofile which is loaded as R launches and is a place where you can place customisations of options to make your R life easier. For instance I have mine set to automatically load workflow helper packages such as `{usethis}` `{testthat}` and `{devtools}`. And set a few options to make R produce more warnings so that it helps me know when something might cause an issue. It's also a good place to set some defaults such as your favourite CRAN mirror so that R doesn't need to ask you when you go to install packages.


```{r, eval=FALSE}
usethis::edit_r_profile()
```


```{r, eval=FALSE}
 # load workflow helper packages if in interactive session
if (interactive()) {
  suppressMessages(require(devtools))
  suppressMessages(require(usethis))
  suppressMessages(require(testthat))
}


# set CRAN
options(repos = c(CRAN = "https://cloud.r-project.org/"))


# warn on partial matches
options(
  warnPartialMatchArgs = TRUE,
  warnPartialMatchDollar = TRUE,
  warnPartialMatchAttr = TRUE
)

# fancy quotes are annoying and lead to
# 'copy + paste' bugs / frustrations
options(useFancyQuotes = FALSE)
```

Don't put analysis modifying options into your .Rprofile e.g. automatic loading of `tidyverse`. It will make the code you write less reproducible because it will start depending on these hidden settings, and it can make debugging harder.

\

#### Keyboard shortcuts

The keyboard shortcut to rule them all: <kbd>Alt</kbd> + <kbd>Shift</kbd> + <kbd>K</kbd>

Other useful shortcuts

Keys | produces
---|---
<kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>M</kbd> | %>%
<kbd>Alt</kbd> + <kbd>-</kbd> | <-
<kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>K</kbd> | knits Rmd
<kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>F10</kbd> | Restart R
<kbd>Ctrl</kbd> + <kbd>Alt</kbd> + <kbd>I</kbd> | Insert code chunk
<kbd>Ctrl</kbd> + <kbd>Shift</kbd> + <kbd>S</kbd> | Source active file
<kbd>Ctrl</kbd> + <kbd>.</kbd> | Go to file/function

Make your own...

Tools -> Modify Keyboard Shortcuts

