---
title: "Multi-plots"
format: html
---

## Setup

```{r}
#| message: false

library(tidyverse)
library(palmerpenguins)

# brings 'penguins' dataset into name space
attach(penguins)
# brings 'penguins_raw' dataset into name space
attach(penguins_raw)
```


## Multi-plots with Patchwork

The {patchwork} package is great for assembling multiple plots into a single plot

Documentation for {patchwork}: https://patchwork.data-imaginist.com/index.html

```{r}
library(patchwork)
```

```{r}
penguin_point_plot <- ggplot(
  data = penguins,
  mapping = aes(x = body_mass_g,
                y = flipper_length_mm,
                colour = island)
) + 
  geom_point() +
  ggtitle("Plot 1")
  

penguin_box_plot <- ggplot(
  data = penguins,
  mapping = aes(x = species,
                y = flipper_length_mm,
                )
) + 
  geom_boxplot() +
  ggtitle("Plot2")
```

Different ways of combining plots

```{r}
penguin_point_plot + penguin_box_plot
```

```{r}
penguin_point_plot / penguin_box_plot
```


Including a table

```{r}
library(gt)
penguin_tab <- penguins %>% 
  group_by(species) %>% 
  summarise(n = n()) %>% 
  gt()  
```

```{r}
penguin_point_plot + penguin_tab
```

### Controlling Layouts


Explain how to control layout (rows/columns, relative sizes, and custom designs) with {patchwork}. Below are short examples using the plots and table defined earlier.

- Use plot_layout() to set number of columns and relative widths/heights.
- Use parentheses to group stacked/side-by-side operators when you want to apply layout options to the combined result.
- Use a design string to create more complex placements (one plot spanning multiple cells).
- Collect legends with guides = "collect" and add a shared title with plot_annotation().


```{r}
# example plots
p1 <- ggplot(penguins, aes(x = body_mass_g, y = flipper_length_mm, colour = species)) +
  geom_point(na.rm = TRUE) + ggtitle("Mass vs Flipper")

p2 <- ggplot(penguins, aes(x = species, y = flipper_length_mm)) +
  geom_boxplot(na.rm = TRUE) + ggtitle("Flipper by species")

p3 <- ggplot(penguins, aes(x = island, y = body_mass_g)) +
  geom_violin(na.rm = TRUE) + ggtitle("Mass by island")

# 1) Two-column layout where p1 takes left column and p2/p3 stack on the right
(p1 | (p2 / p3)) +
  plot_layout(widths = c(2, 1)) +
  plot_annotation(title = "p1 large left, p2 & p3 stacked right")

# 2) Grid layout with relative widths/heights and a shared legend collected
(p1 + p2 + p3) +
  plot_layout(ncol = 2, widths = c(2, 1), heights = c(1, 1), guides = "collect") + #&
  # theme(legend.position = "bottom") +
  plot_annotation(tag_levels = "A")

# 3) Complex placement using a design string (A spans 3 cells in row 1)
design <- "
AAA
B#C
"
p1 + p2 + p3 + plot_layout(design = design) +
  plot_annotation(title = "Design string layout")

```

Notes: use widths/heights for relative sizing, guides = "collect" to share legends, and a design string for precise cell placement.
