---
title: "02-wrangle"
format: html
---

## dplyr refresher with palmerpenguins

This short refresher demonstrates select, filter, mutate, group_by, and summarise using the palmerpenguins dataset.

```{r}
library(tidyverse)
library(palmerpenguins)

# minimal clean dataset for examples
penguins_clean <- penguins %>%
  select(species, island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex) %>%
  drop_na()
```

select — keep specific columns
```{r}
penguins_clean %>%
  select(species, island, body_mass_g) %>%
  slice_head(n = 6)
```

filter — keep rows matching conditions
```{r}
# filter by species and body mass > 4000 g
penguins_clean %>%
  filter(species == "Adelie", body_mass_g > 4000) %>%
  slice_head(n = 6)
```

mutate — add or transform columns
```{r}
# add body mass in kg and bill ratio
penguins_clean %>%
  mutate(
    body_mass_kg = body_mass_g / 1000,
    bill_ratio = bill_length_mm / bill_depth_mm
  ) %>%
  slice_head(n = 6)
```

group_by + summarise — aggregate by groups
```{r}
# mean body mass and median flipper length per species
penguins_clean %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    median_flipper = median(flipper_length_mm, na.rm = TRUE)
  ) %>%
  arrange(desc(mean_body_mass))
```
```

## Using across() and where()

Use across() to apply the same transformation or summary to multiple columns; use where() to select columns by predicate (e.g., is.numeric, is.character).

```{r}
# 1) mutate with across + where: convert numeric measurements to centimeters
penguins_clean %>%
  mutate(across(where(is.numeric), ~ . / 10, .names = "{.col}_cm")) %>%
  select(species, ends_with("_cm")) %>%
  slice_head(n = 6)
```

```{r}
# 2) mutate with across on character columns: normalize text
penguins_clean %>%
  mutate(across(where(is.character), ~ str_to_title(.))) %>%
  slice_head(n = 6)
```

```{r}
# 3) summarise with across: compute mean and sd for all numeric columns per species
penguins_clean %>%
  group_by(species) %>%
  summarise(across(where(is.numeric),
                   list(mean = ~ mean(.x, na.rm = TRUE),
                        sd   = ~ sd(.x, na.rm = TRUE)),
                   .names = "{.col}_{.fn}"))
```

```{r}
# 4) summarise with across + select helpers: mean of bill measurements
penguins_clean %>%
  group_by(species) %>%
  summarise(across(starts_with("bill_"), mean, na.rm = TRUE, .names = "mean_{.col}"))
```

## .names examples

# 1) When you supply a named list of functions, {.fn} uses the names you provided


```{r}
penguins_clean %>%
    group_by(species) %>%
    summarise(
        across(where(is.numeric),
                     list(avg = ~ mean(.x, na.rm = TRUE),
                                sdev = ~ sd(.x, na.rm = TRUE)),
                     .names = "{.col}_{.fn}")   # e.g. bill_length_mm_avg, bill_length_mm_sdev
    )
```

2) For a single function {.fn} defaults to the function name unless you hardcode a label

```{r}
penguins_clean %>%
    group_by(species) %>%
    summarise(
        across(starts_with("bill_"),
                     mean, na.rm = TRUE,
                     .names = "mean_{.col}")   # e.g. mean_bill_length_mm
    )
```


3) Use .names to add units/formatting when mutating multiple columns

```{r}
penguins_clean %>%
    mutate(across(starts_with("bill_"),
                                ~ .x / 10,
                                .names = "{.col}_cm")) %>%   # e.g. bill_length_mm_cm
    select(species, ends_with("_cm")) %>%
    slice_head(n = 6)
```

Notes
- across() replaced many uses of mutate_at()/summarise_at() and is the recommended tidyverse approach.
- where() is handy when you want to target all numeric or character columns without naming them explicitly.
- Use .names in across() to control output column naming and avoid collisions.
