
---
title: "03-tables"
format: html
---

## Quick tutorial: gt with the palmerpenguins dataset

This short tutorial shows common gt workflows: build a basic table, format numbers and labels, add color scales, and produce a grouped summary.


## Setup

Install and load packages (only once):

```{r}
# install.packages(c("gt", "palmerpenguins", "dplyr", "scales"))
library(tidyverse)
library(gt)
library(palmerpenguins)
library(scales)
```

1) Basic table
Select a handful of columns and render a simple gt table.

```{r}
penguins_clean <- penguins %>%
  select(species, island, bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g, sex) %>%
  drop_na()

penguins_clean %>%
  slice_head(n = 10) %>%
  gt()
```


2) Format numbers and labels
Use gt formatting helpers to control decimal places and column labels.


```{r}
penguins_clean %>%
  slice_head(n = 10) %>%
  gt() %>%
  cols_label(
    species = "Species",
    island = "Island",
    bill_length_mm = "Bill (mm)",
    bill_depth_mm  = "Bill depth (mm)",
    flipper_length_mm = "Flipper (mm)",
    body_mass_g = "Body mass (g)",
    sex = "Sex"
  ) %>%
  fmt_number(
    columns = c("bill_length_mm", "bill_depth_mm", "flipper_length_mm"),
    decimals = 1
  ) %>%
  fmt_number(
    columns = c("body_mass_g"),
    decimals = 0
  )
```


3) Add a color scale
Highlight body mass with a color ramp to make patterns visible.
```{r}
body_range <- range(penguins_clean$body_mass_g, na.rm = TRUE)

penguins_clean %>%
  slice_head(n = 10) %>%
  gt() %>%
  cols_label(body_mass_g = "Body mass (g)") %>%
  data_color(
    columns = c("body_mass_g"),
    colors = col_numeric(
      palette = c("lightyellow", "orange", "red"),
      domain = body_range
    )
  ) %>%
  fmt_number(columns = vars(body_mass_g), decimals = 0)
```


Using scales to create the colour mapper
The function col_numeric() used above comes from the scales package (scales::col_numeric). It builds and returns a function that maps numeric values to colors; you pass that function to gt::data_color so gt can colour cells based on the numeric values. The two important arguments are palette (the colour stops — use hex codes or a perceptually uniform palette like viridis) and domain (the numeric range to map).

Tip: construct the mapping function explicitly so you can inspect or reuse it, and prefer explicit domains (range(..., na.rm = TRUE)) and hex colours for reproducibility and accessibility.

```{r}
# create a reusable colour-mapping function (example using viridis)
library(viridisLite)   # lightweight dependency for perceptual colours

color_fn <- scales::col_numeric(
  palette = viridisLite::viridis(5),                       # perceptually uniform palette
  domain  = range(penguins_clean$body_mass_g, na.rm = TRUE) # explicit numeric domain
)

# inspect output for a few values
color_fn(c(2700, 3500, 5000))

# use the function with gt::data_color
penguins_clean %>%
  slice_head(n = 10) %>%
  gt() %>%
  cols_label(body_mass_g = "Body mass (g)") %>%
  data_color(columns = c("body_mass_g"), colors = color_fn) %>%
  fmt_number(columns = vars(body_mass_g), decimals = 0)
```
For other needs, scales also provides helpers like col_bin(), col_quantile(), and col_factor() to create binned, quantile-based, or categorical colour mappings. ``````




4) Grouped summary (per species)
Create a summary table grouped by species: mean body mass and median flipper length.


```{r}
summary_tbl <- penguins_clean %>%
  group_by(species) %>%
  summarise(
    n = n(),
    mean_body_mass = mean(body_mass_g, na.rm = TRUE),
    median_flipper = median(flipper_length_mm, na.rm = TRUE)
  ) %>%
  ungroup()

summary_tbl %>%
  gt(rowname_col = "species") %>%
  cols_label(
    n = "Count",
    mean_body_mass = "Mean body mass (g)",
    median_flipper = "Median flipper (mm)"
  ) %>%
  fmt_number(columns = vars(mean_body_mass, median_flipper), decimals = 1)
```


5) Grouped details plus summary rows
Show species groups with a grand summary row.


```{r}
penguins_clean %>%
  group_by(species) %>%
  slice_sample(n = 6) %>%    # small sample per species for display
  ungroup() %>%
  gt(groupname_col = "species") %>%
  summary_rows(
    groups = TRUE,
    columns = c("body_mass_g"),
    fns = list(
      "avg (g)" = ~mean(., na.rm = TRUE),
      "min (g)" = ~min(., na.rm = TRUE),
      "max (g)" = ~max(., na.rm = TRUE)
    )
  ) %>%
  grand_summary_rows(
    columns = c("body_mass_g"),
    fns = list("Overall avg (g)" = ~mean(., na.rm = TRUE))
  ) %>%
  fmt_number(columns = vars(body_mass_g), decimals = 0)
```
